Formula translation in \changed{Blitz++, NumPy} and modern Fortran: a~case~study~of~the~language~choice tradeoffs

### 2014-07-01 paper/doc/grid.tex
adding grid.tex

\documentclass[twocolumn]{article}
\usepackage{authblk,latexsym,amsmath,float}
\usepackage[runin]{abstract}
\usepackage[colorlinks=true,allcolors=blue]{hyperref}
\usepackage[hmargin=.55in,vmargin=.75in]{geometry}
\usepackage[numbers]{natbib}

% do-nothing commands
\newcommand{\changed}[1]{#1}
\newcommand{\removed}[1]{#1}

% input encoding assuring proper handling of Polish characters (surnames!)
\usepackage[utf8]{inputenc}
\usepackage{fontenc}

\usepackage{type1cm}
\usepackage{relsize}

\input{formulae_preamble}

% for Arakawa-C figure
\usepackage{tikz}
\usetikzlibrary{calc}

% for conpact itemize etc
\usepackage{enumitem}
\setitemize{noitemsep,itemsep=1pt,topsep=1pt,parsep=0pt,partopsep=0pt}

% for compact section headers
\usepackage[compact]{titlesec}

% for source code listings
\usepackage{fancyvrb}
\fvset{
  frame=single,
  fontsize=\relscale{.66},
  fontfamily=courier,
  framerule=.2mm,
  framesep=.7mm,
  numbers=left,
  xleftmargin=4mm,
  numbersep=.25mm,
  firstnumber=last
}
\input{pygments}

\newcounter{lstnocpp}
\newcounter{lstnopyt}
\newcounter{lstnofor}

\setcounter{lstnocpp}{-1}
\setcounter{lstnopyt}{-1}
\setcounter{lstnofor}{-1}

\newcounter{linenocpp}
\newcounter{linenopyt}
\newcounter{linenofor}

\setcounter{linenocpp}{0}
\setcounter{linenopyt}{0}
\setcounter{linenofor}{0}

\newcommand*\FancyVerbStartString{}
\newcommand*\FancyVerbStopString{}

\newcommand{\codecpp}[4]{%
  \addtocounter{lstnocpp}{1}%
  \renewcommand*\FancyVerbStartString{\PY{c+c1}{//#2}}%
  \renewcommand*\FancyVerbStopString{\PY{c+c1}{//#3}}%
  \setcounter{FancyVerbLine}{\thelinenocpp}%
  \fvset{label={listing~C.\thelstnocpp~(C++)},rulecolor=\color{black},stepnumber=#4}%
  \input{#1}%
  \setcounter{linenocpp}{\value{FancyVerbLine}}%
}
\newcommand{\codepyt}[4]{%
  \addtocounter{lstnopyt}{1}%
  \renewcommand*\FancyVerbStartString{\PY{c}{\PYZsh{}#2}}
  \renewcommand*\FancyVerbStopString{\PY{c}{\PYZsh{}#3}}
  \setcounter{FancyVerbLine}{\thelinenopyt}%
  \fvset{label={listing~P.\thelstnopyt~(Python)},rulecolor=\color{blue},stepnumber=#4}%
  \input{#1}%
  \setcounter{linenopyt}{\value{FancyVerbLine}}%
}
\newcommand{\codefor}[4]{%
  \addtocounter{lstnofor}{1}%
  \renewcommand*\FancyVerbStartString{\PY{c}{!#2}}
  \renewcommand*\FancyVerbStopString{\PY{c}{!#3}}
  \setcounter{FancyVerbLine}{\thelinenofor}%
  \fvset{label={listing~F.\thelstnofor~(Fortran)},rulecolor=\color{red},stepnumber=#4}%
  \input{#1}%
  \setcounter{linenofor}{\value{FancyVerbLine}}%
}

\newcommand{\prog}[1]{{\rm\bf#1}}

\begin{document}

    \title{
      Formula translation in \changed{Blitz++, NumPy} and modern Fortran: a~case~study~of~the~language~choice tradeoffs
    }

    \date{}

    \author[1]{Sylwester Arabas}%\footnote{Correspondence to: S. Arabas -- sarabas@igf.fuw.edu.pl -- University of Warsaw, Pasteura 7, 02-093 Warszawa, Poland}}
    \author[1]{Dorota Jarecka}
    \author[1]{Anna Jaruga}
    \author[2]{Maciej Fijałkowski}

    \affil[1]{Institute of Geophysics, Faculty of Physics, University of Warsaw}
    \affil[2]{PyPy Team}

    \twocolumn[
      \maketitle

      \begin{onecolabstract}
	  Three object-oriented implementations of a prototype solver of the advection equation are introduced.
	  The presented programs are based on Blitz++ (C++), NumPy (Python), and Fortran's built-in array containers.
	  The solvers constitute implementations of the Multidimensional Positive-Definite 
	    Advective Transport Algorithm (MPDATA).
	  The introduced codes serve as examples for how the application of 
	    object-oriented programming (OOP) techniques and new language constructs from
	    C++11 and Fortran 2008 allow to reproduce the mathematical notation 
	    used in the literature within the program code.
	  A discussion on the tradeoffs of the programming language choice is presented.
	  The main angles of comparison are code brevity and syntax clarity
	    (and hence maintainability and auditability) as well as performance.
	  All performance tests are carried out using free and open-source compilers.
	  In the case of Python, a significant performance gain is observed when switching from the standard 
	    interpreter (CPython) to the PyPy implementation of Python.
	  Entire source code of all three implementations is embedded in the text and is licensed
	    under the terms of the GNU GPL license.

	  ~\\
	  {\bf Keywords:} object-oriented programming, advection equation, MPDATA, C++, Fortran, Python
	  ~\\
      \end{onecolabstract}
  ]

  \tableofcontents
  \newpage

  \section{Introduction}

  Object oriented programming (OOP) {\em ``has become recognised as the almost unique successful 
    paradigm for creating complex software''} \citep[][Sec.~1.3]{Press_et_al_2007}.
  It is intriguing that, while the quoted statement comes from the very book subtitled 
   {\em The Art of Scientific Computing}, hardly any (if not none) of the currently operational 
    weather and climate prediction systems - flagship examples of complex scientific software - 
    make extensive use of OOP techniques\footnote{
  Fortran has been the language of choice in oceanic \citep{Griffies_et_al_2000}, 
    weather-prediction \citep{Sundberg_2009} and Earth system \citep{Legutke_2012} modelling, 
    and none of its 20\textsuperscript{th}-century editions were object-oriented languages \citep[see e.g.][for discussion]{Norton_et_al_2007}}.

  Application of OOP techniques in development of numerical modelling software may help to:
  \begin{enumerate}[label=(\roman*), leftmargin=*, widest=ii]
    \item{maintain modularity and separation of program logic layers (e.g. separation of
      numerical algorithms, parallelisation mechanisms, data input/output, error handling and
      the description of physical processes); and}
    \item{{\bf shorten and simplify the source code and improve its readability by reproducing within 
      the program logic the mathematical notation used in the literature}.}
  \end{enumerate}
  The first application is attainable, yet arguably cumbersome, with procedural programming.
  The latter, virtually impossible to obtain with procedural programming, is the focus of this paper.
  The importance of reproducing the mathematical notation in the code lays primarily
    in the fact that code readability and brevity 
    %\citep[and not the language choice itself,][]{Seeley_2004}
    significantly contribute to code maintainability \citep{Wilson_et_al_2012}.

  The key aim of this paper is to show how OOP techniques can be used to faithfully
    reproduce within the code what can be referred to as {\em blackboard abstractions} \citep{Rouson_et_al_2012}.
  \changed{These may relate to several levels of mathematical abstraction. 
    Object-oriented logic can be used to make the code resemble analytical formul\ae~\citep[e.g.][]{Weller_et_al_1998}
    and/or numerical algorithms, the latter being exemplified in this paper.
    }
  For this purpose, a sample implementation of a numerical scheme for solving
    the advection equation is introduced in C++, Python and modern Fortran --
    OOP languages commonly used in scientific computing \citep[see e.g.][chapt.~8]{Einarsson_2005}. 
  Presented implementations and the results of benchmark tests
    provide a basis for discussion on the tradeoffs of programming language choice.
  The discussion concerns in principle the development of finite-difference solvers for 
    partial differential equations, but is likely applicable to some extent to the scientific programming in general.

  All three programs include an equally structured implementation of the two-dimensional version of the 
    Multidimensional Positive Definite Advective Transport Algorithm \citep[MPDATA;][]{Smolarkiewicz_1984}.
  MPDATA is an example of a numerical procedure used in weather, climate, ocean and solar simulation systems
    \citep[e.g.][respectively]{Ziemianski_et_al_2011,Abiodun_et_al_2011,Ezer_et_al_2002,Charbonneau_and_Smolarkiewicz_2013}.
  The basic MPDATA scheme presented herein is complex enough to contain
    a wide range of mathematical abstractions that can be represented using OOP constructs, 
    yet it is simple enough to allow inclusion of the entire source code within the paper text.
  All relevant MPDATA formul\ae~are given in the text alongside corresponding code fragments
    allowing comparison of the relevant syntax with the mathematical notation.
  These formul\ae~are presented without derivation or detailed discussion
    (see~\citep{Smolarkiewicz_2006} for a recent review of MPDATA-based techniques including an introductory description 
    of the algorithm and an exhaustive list of references).

  The paper is structured as follows.
  In section~\ref{sec:impl} we introduce the ``formula translation'' part of the three implementations
    briefly describing the algorithm itself and
    discussing where and how the OOP techniques were applied in its implementation.
  The remaining part of the implementations -- the solver logic -- is
    presented in appendix~\ref{app:solver}.
  A usage example is given in appendix~\ref{app:usage}
  Section~\ref{sec:perf} covers performance evaluation of the three implementations.
  Section~\ref{sec:tradeoffs} covers discussion of the tradeoffs of the programming language choice.
  Section~\ref{sec:concl} closes the article with a brief summary.

  The entire code is licensed under the terms of the GNU General Public License version 3 \citep{GPLv3}.
  All listings include line numbers printed to the left of the source code, with separate numbering for
    C++ (listings prefixed with C, black frame),
  \codecpp{code-cpp-listings.hpp}{listing00}{listing01}{1}
    Python (listings prefixed with P, blue frame) and
  \codepyt{code-pyt-listings.py}{listing00}{listing01}{1}
    Fortran (listings prefixed with F, red frame).
  \codefor{code-for-listings.f}{listing00}{listing01}{1}
  Programming language constructs when inlined in the text are 
    typeset in bold, e.g. \prog{GOTO 2}.

  \section{Implementation of the formul\ae}\label{sec:impl}

  Double precision floating-point format is used in all three implementations.
  The codes begin with the following definitions:
  \codecpp{code-cpp-listings.hpp}{listing01}{listing02}{1}
  \codepyt{code-pyt-listings.py}{listing01}{listing02}{1}
  \codefor{code-for-listings.f}{listing01}{listing02}{1}
  which provide a convenient way of switching to different precision\footnote{Fortran's
    \prog{selected\_real\_kind()}" intrinsic function may be used instead to improve portability.}.

  All codes are structured in a way allowing compilation of the code 
    in exactly the same order as presented in the text within one source file,
    hence every Fortran listing contains definition of a separate module.

  The language syntax and OOP nomenclature are used without introduction in the paper,
    for an overview of OOP in context of C++, Python and Fortran, consult for example
    \citep[][Part~III]{Stroustrup_2013}, \citep[][Chapter~5]{Pilgrim_2004} and
    \citep[][Chapter~11]{Markus_2012}, respectively.

  \subsection{Array containers}\label{sec:array}

  MPDATA is, in its most basic form presented herein, a solver for systems of advection equations of the following form:

  \begin{equation}\label{eq:adv}
    \input{formulae_adv}
  \end{equation}
  that describe evolution of a scalar field $\psi$ transported by the fluid flow with velocity $\vec{v}$.
  Solution of equation (\ref{eq:adv}) using MPDATA implies discretisation onto a grid of~the~$\psi$~and~the~Courant 
    number $\vec{C}=\vec{v}\cdot\frac{\Delta t}{\Delta x}$ fields, where $\Delta t$ is the 
    solver timestep and $\Delta x$ is the grid spacing.

  Presented C++ implementation of MPDATA is built upon the Blitz++ 
    library\footnote{Blitz++ is a C++ class library for scientific computing which uses the expression templates technique to achieve high performance, see \url{http://sf.net/projects/blitz/}}.
  Blitz offers object-oriented representation of n-dimensional arrays,
    and array-valued mathematical expressions.
  In particular, it offers loop-free notation for array arithmetics
    that does not incur creation of intermediate temporary objects.
  Blitz++ is a header-only library\footnote{Blitz++ requires linking with \prog{libblitz} if debugging mode is used} -- to use it, it is enough to include the appropriate header file,
    and optionally expose the required classes to the present namespace:
  \codecpp{code-cpp-listings.hpp}{listing02}{listing03}{1}
  Here \prog{arr\_t}, \prog{rng\_t} and \prog{idx\_t} serve as alias identifiers 
    and are introduced in order to shorten the code.

  The power of Blitz++ comes from the ability to express array expressions as objects.
  In particular, it is possible to define a function that returns an array expression;
    i.e. not the resultant array, but an object representing a ``recipe'' defining the operations
    to be performed on the arguments.
  As a consequence, the return types of such functions become unintelligible.
  Luckily, the \prog{auto} return type declaration from the C++11 standard allows to simplify the code significantly,
    even more if used through the following preprocessor macro:
  \codecpp{code-cpp-listings.hpp}{listing03}{listing04}{1}
  The call to \prog{blitz::safeToReturn()} function is included in order to ensure that
    all arrays involved in the expression being returned continue to exist in the
    caller scope.
  For example, definition of a function returning its array-valued argument doubled, reads:
    \prog{auto~f(arr\_t~x)~return\_macro(2*x)}.
  This is the only preprocessor macro defined herein.

  For the Python implementation of MPDATA, the NumPy\footnote{NumPy is a Python package
    for scientific computing offering support for multi-dimensional arrays and a library
    of numerical algorithms, see \url{http://numpy.org/}} package is used.
  In order to make the code compatible with both the standard CPython
    as well as the alternative PyPy implementation of Python \citep[][]{Bolz_et_al_2011},
    the Python code includes the following sequence of \prog{import} statements: 
  \codepyt{code-pyt-listings.py}{listing02}{listing03}{1}
  First, the PyPy's built-in NumPy implementation named \prog{numpypy} is imported if applicable (i.e. if running PyPy), 
    and the lazy evaluation mode is turned on through the \prog{set\_invalidation(False)} call.
  PyPy's lazy evaluation obtained with the help of a just-in-time compiler enables to achieve
    an analogous to Blitz++ temporary-array-free handling of array-valued expressions 
    (see discussion in section~\ref{sec:perf}).
  Second, to match the settings of C++ and Fortran compilers used herein, the NumPy package is instructed 
    to ignore any floating-point errors, if such an option
    is available in the interpreter\footnote{\prog{numpy.seterr()} is not supported in PyPy as of version 1.9}.
  The above lines conclude all code modifications that needed to be added in order to run
    the code with PyPy.

  Among the three considered languages only Fortran is equipped with built-in
    array handling facilities of practical use in high-performance computing. 
  Therefore, there is no need for using an external package as with C++ and Python.  
  Fortran array-handling features are not object-oriented, though \changed{(e.g. it is impossible
    to overload array operators or to provide constructor-like initialisation logic)}.

  \subsection{Containers for sequences of arrays}\label{sec:sequence}

  As discussed above, discretisation in space of the scalar field $\psi(x,y)$ into its $\psi_{[i,j]}$ 
    grid representation requires floating-point array containers.
  In turn, discretisation in time requires a container class for storing
    sequences of such arrays, i.e. \{$\psi^{[n]}$, $\psi^{[n+1]}$\}.
  Similarly the components of the vector field $\vec{C}$ are in fact a \{$C^{[x]}$, $C^{[y]}$\} 
    array sequence.

  Using an additional array dimension to represent the sequence elements is not considered for two reasons.
  First, the $C^{[x]}$ and $C^{[y]}$ arrays constituting the sequence have different sizes
    (see discussion of the Arakawa-C grid in section~\ref{sec:grid}).
  Second, the order of dimensions would need to be different for different languages to assure that
    the contiguous dimension is used for one of the space dimensions and not for time levels.

  In the C++ implementation, the Boost\footnote{
    Boost is a free and open-source collection of peer-reviewed C++ libraries available at \url{http://www.boost.org/}.
    Several parts of Boost have been integrated into or inspired new additions to the C++ standard.
  } \prog{ptr\_vector} class is used to represent sequences of Blitz++ arrays 
    and at the same time to handle automatic freeing of dynamically allocated memory.
  The \prog{ptr\_vector} class is further customised by defining a derived structure which element-access \prog{[~]} 
    operator is overloaded with a modulo variant:
  \codecpp{code-cpp-listings.hpp}{listing04}{listing05}{1}
  Consequently the last element of any such sequence may be accessed at index \prog{-1}, the last but one at \prog{-2}, 
    and so on.

  In the Python implementation the built-in \prog{tuple} type is used to store sequences of NumPy arrays.
  Employment of negative indices for handling from-the-end addressing of elements
    is a built-in feature of all sequence containers in Python.

  Fortran does not feature any built-in sequence container capable of storing arrays, 
    hence a custom \prog{arrvec\_t} type is introduced:
  \codefor{code-for-listings.f}{listing02}{listing03}{1}
  The \prog{arr\_t} type is defined solely for the purpose of overcoming the limitation 
    of lack of an array-of-arrays construct, and its only member field is a two-dimensional array.
  An array of \prog{arr\_t} is used hereinafter as a container for sequences of arrays.

  The \prog{arrptr\_t} type is defined solely for the purpose of overcoming Fortran's limitation
    of not supporting allocatables of pointers.
  \prog{arrptr\_t}'s single member field is a pointer to an instance of \prog{arr\_t}.
  Creating an allocatable of \prog{arrptr\_t}, instead of 
    a multi-element pointer of \prog{arr\_t}, ensures automatic memory deallocation.

  Type \prog{arrptr\_t} is used to implement the from-the-end addressing of elements in \prog{arrvec\_t}.
  The array data is stored in the \prog{arrs} member field (of type \prog{arr\_t}).
  The \prog{at} member field (of type \prog{arrptr\_t}) stores pointers to the elements of \prog{arrs}. 
  \prog{at} has double the length of \prog{arrs} and is initialised in a cyclic manner so that
    the \prog{-1} element of \prog{at} points to the last element of \prog{arrs}, and so on.
  Assuming \prog{psi} is an instance of \prog{arrptr\_t}, the \prog{(i,j)} element of the \prog{n}-th array in \prog{psi}
    may be accessed with\\ \prog{psi\%at( n )\%p\%a( i, j )}.

  The \prog{ctor(n)} method initialises the container for a given number of elements \prog{n}.
  The \prog{init(n,i,j)} method initialises the \prog{n}-th element of the container with 
    a newly allocated 2D array spanning indices \prog{i(1)}:\prog{i(2)}, 
    and \prog{j(1)}:\prog{j(2)} in the first, and last dimensions respectively\footnote{In Fortran, when an array
    is passed as a function argument its base is locally set to unity, regardless of the setting
    at the caller scope.}.

  \subsection{Staggered grid}\label{sec:grid}

  \begin{figure}[h!]
  \center
  \begin{tikzpicture}
    \coordinate (Origin)   at (0,0);
    \coordinate (XAxisMin) at (-2.5,0);
    \coordinate (XAxisMax) at (3,0);
    \coordinate (YAxisMin) at (0,-2.5);
    \coordinate (YAxisMax) at (0,3);
    \draw [thin, gray,-latex] (XAxisMin) -- (XAxisMax);% Draw x axis
    \draw [thin, gray,-latex] (YAxisMin) -- (YAxisMax);% Draw y axis

@@

    \draw [ultra thick,-latex,red] (0,.8) -- (0,1.2) node [above] {};
    \draw [ultra thick,-latex,red] (0,-1.2) -- (0,-.8) node [below right] {$C^{[y]}_{[i,j\mhlf]}$};
    \draw [ultra thick,-latex,red] (-2,.8) -- (-2,1.2) node [above] {};
    \draw [ultra thick,-latex,red] (-2,-1.2) -- (-2,-.8) node [above] {};
    \draw [ultra thick,-latex,red] (2,.8) -- (2,1.2) node [above] {};
    \draw [ultra thick,-latex,red] (2,-1.2) -- (2,-.8) node [above] {};
  \end{tikzpicture}
  \caption{\label{fig:grid}
    A schematic of the Arakawa-C grid.
  }
  \end{figure}
  The so-called Arakawa-C staggered grid \citep{Arakawa_and_Lamb_1977} depicted in Figure~\ref{fig:grid}
    is a natural choice for MPDATA.
  As a consequence, the discretised representations of the $\psi$ scalar field, and each 
    component of the ${\vec{C}=\vec{v}\cdot\frac{\Delta t}{\Delta x}}$ vector field 
    in eq.~(\ref{eq:adv}) are defined over different grid point locations.
  In mathematical notation this can be indicated by usage of fractional indices, e.g.
    $C^{[x]}_{[i\mhlf,j]}$, $C^{[x]}_{[i+\phlf,j]}$, $C^{[y]}_{[i,j\mhlf]}$ and $C^{[y]}_{[i,j+\phlf]}$
    to depict the grid values of the ${\vec{C}}$ vector components surrounding $\psi_{[i,j]}$.
  However, fractional indexing does not have a built-in counterpart in any of the
    employed programming languages.
  A desired syntax would translate \prog{$i-\phlf$} to \prog{$i-1$} and
    \prog{$i+\phlf$} to \prog{$i$}.
  OOP offers a convenient way to implement such notation
    by overloading the \prog{+} and \prog{-} operators for objects representing array indices.

  In the C++ implementation, first a global instance \prog{h} of an empty structure 
    \prog{hlf\_t} is defined, and then the plus and minus operators for \prog{hlf\_t} and \prog{rng\_t} are overloaded:
  \codecpp{code-cpp-listings.hpp}{listing05}{listing06}{1}
  This way, the arrays representing vector field components can be indexed using
    \prog{(i+h,j)}, \prog{(i-h,j)} etc. where \prog{h}~represents the half.

  In NumPy, in order to prevent copying of array data during slicing, one needs to operate on the
    so-called array views.
  Array views are obtained when indexing the arrays with objects of the Python's
    built-it \prog{slice} type (or tuples of such objects in case of multi-dimensional arrays).
  Python forbids overloading of operators of built-in types such as \prog{slices}, 
    and does not define addition/subtraction operators for \prog{slice} and \prog{int} pairs.
  Consequently, a custom logic has to be defined not only for fractional indexing,
    but also for shifting the slices by integer intervals ($i\pm1$).
  It is implemented here by declaring a \prog{shift} class with the adequate operator overloads:
  \codepyt{code-pyt-listings.py}{listing03}{listing04}{1}
    and two instances of it to represent unity and half
    in expressions like \prog{i+one}, \prog{i+hlf}, where \prog{i} is an instance of \prog{slice}%
    \footnote{\label{fnt:slice}One could argue that not using an own implementation of a slice-representing class
    in NumPy is a design flaw -- being able to modify behaviour of a hypothetical numpy.slice class 
    through inheritance would allow to implement the same behaviour as obtained in listing P.3 without the need to represent 
    the unity as a separate object}:
  \codepyt{code-pyt-listings.py}{listing04}{listing05}{1}

  In the Fortran implementation fractional array indexing is obtained through
    definition and instantiation of an object representing the half, and having appropriate
    operator overloads:
  \codefor{code-for-listings.f}{listing03}{listing04}{1}

  \subsection{Array index permutations}\label{sec:pi}
  Hereinafter, the $\pi_{a,b}^{d}$ symbol is used to denote a  
    cyclic permutation of an order $d$ of a set $\{a,b\}$.
  It is used to generalise the MPDATA formul\ae~into multiple dimensions
    using the following notation:

  \begin{equation*}\label{eq:pi}
    \input{formulae_pi}
  \end{equation*}

  Blitz++ ships with the \prog{RectDomain} class (aliased here as \prog{idx\_t}) 
    for specifying array ranges in multiple dimensions.
  The $\pi$ permutation is implemented in C++ as a function \prog{pi()}
    returning an instance of \prog{idx\_t}.
  In order to ensure compile-time evaluation, the permutation order is passed 
    via the template parameter \prog{d}
    (note the different order of \prog{i} and \prog{j} arguments in the two template specialisations):
  \codecpp{code-cpp-listings.hpp}{listing06}{listing07}{1}

  NumPy uses tuples of slices for addressing multi-dimensional array with a
    single object.
  Therefore, the following definition of function \prog{pi()} suffices to represent~$\pi$:
  \codepyt{code-pyt-listings.py}{listing05}{listing06}{1}

  Fortran does not feature an analogous mechanism for specifying array ranges in multiple
    dimensions using a single entity.
  As a workaround, in the Fortran implementation, \prog{pi()} returns a pointer to the array elements specified
    by \prog{i} and \prog{j} interpreted as (i,j) or
    (j,i) depending on the value of the argument \prog{d}.
  In addition to \prog{pi()}, a helper \prog{span()} function returning the 
    length of one of the vectors passed as argument is defined:
  \codefor{code-for-listings.f}{listing04}{listing05}{1}
  The \prog{span()} function is used to shorten the declarations of arrays to be returned 
    from functions in the Fortran implementation (see listings F.7 and F.12--F.15).

  It is worth noting here that the C++ implementation of \prog{pi()} is
    branchless thanks to employment of template specialisation.
  With Fortran one needs to rely on compiler optimisations to eliminate the conditional expression
    within \prog{pi()} that depends on value of \prog{d} which is always known 
    at compile time.

  \subsection{Donor-cell formul\ae}\label{sec:donor}

  MPDATA is an iterative algorithm in which each iteration takes the form of the 
    so-called donor-cell formula (which itself is a first-order advection scheme).

  MPDATA and donor-cell are explicit forward-in-time algorithms -- they allow to predict $\psi^{[n+1]}$ as a
    function of $\psi^{[n]}$ where $n$ and $n+1$ denote two adjacent time levels.
  The donor-cell scheme may be written as \citep[eq. 2 in][]{Smolarkiewicz_1984}:

  \begin{equation}\label{eq:donor}
    \input{formulae_donor}
  \end{equation}
  where $N$ is the number of dimensions, 
    and F is the so-called flux function \citep[][eq.~3]{Smolarkiewicz_1984}:

  \begin{equation}\label{eq:donor:F}
    \input{formulae_donor-F}
  \end{equation}

  \noindent The flux function takes the following form in C++:
  \codecpp{code-cpp-listings.hpp}{listing07}{listing08}{1}
  Equation~\ref{eq:donor} is split into the terms under the summation 
    (effectively the 1-dimensional donor-cell formula):
  \codecpp{code-cpp-listings.hpp}{listing08}{listing09}{1}
  and the actual two-dimensional donor-cell formula:
  \codecpp{code-cpp-listings.hpp}{listing09}{listing10}{1}
  The same formul\ae~are expressed as follows in Python:
  \codepyt{code-pyt-listings.py}{listing06}{listing07}{1}
  \codepyt{code-pyt-listings.py}{listing07}{listing08}{1}
  \codepyt{code-pyt-listings.py}{listing08}{listing09}{1}
  The Fortran counterparts are: 
  \codefor{code-for-listings.f}{listing05}{listing06}{1}
  \codefor{code-for-listings.f}{listing06}{listing07}{1}
  \codefor{code-for-listings.f}{listing07}{listing08}{1}
  \codefor{code-for-listings.f}{listing08}{listing09}{1}
  \codefor{code-for-listings.f}{listing09}{listing10}{1}

  The brevity of the code in the above listings as well as its similarity
    to the mathematical notation is the main point of this paper.
  The ``formula translation'' features include:
  \begin{itemize}
    \item{loop-free notation;}
    \item{array-valued functions enabling reuse of sub-expressions;}
    \item{fractional indexing obtained with the help of operator overloading;}
    \item{dimension-independent indexing with the help of \prog{pi} permutation functions.}
  \end{itemize}
  The same features are applied to translation of more complex formul\ae~in the
    following section.

  \subsection{MPDATA formul\ae}\label{sec:mpdata}

  MPDATA introduces corrective steps to the algorithm defined by equation~\ref{eq:donor} and \ref{eq:donor:F}.
  Each corrective step is a donor-cell step (eq.~\ref{eq:donor}) with the Courant number
    fields corresponding to the MPDATA antidiffusive velocities of the following 
    form \citep[eqs~13,~14~in][]{Smolarkiewicz_1984}:

  \begin{equation}\label{eq:antidiff1}
    \input{formulae_antidiff1}
  \end{equation}
  where $\psi$ and $C$ represent values from the previous iteration and where:

  \begin{equation}\label{eq:antidiff2}
    \input{formulae_antidiff2}
  \end{equation}
  For positive-definite $\psi$, the $A$ and $B$ terms take the 
  following form\footnote{
    Since $\psi\ge0$, $|A|\le1$ and $|B|\le1$. 
    See \citet[Sec. 4.2]{Smolarkiewicz_2006} for description of adaptation of the 
    formul\ae~for advection of fields of variable sign
  }:

  \begin{equation}\label{eq:A}
    \input{formulae_A}
  \end{equation}
  \begin{equation}\label{eq:B}
    \input{formulae_B}
  \end{equation}

  If the (positive-defined) denominator in equations \ref{eq:A} or \ref{eq:B} 
    equals zero for a given {\em i} and {\em j},
    the corresponding $A_{[i,j]}$ and $B_{[i,j]}$ are set to zero what may be conveniently 
    represented with the \prog{where} construct in all three considered languages:
  \codecpp{code-cpp-listings.hpp}{listing10}{listing11}{1}
  \codepyt{code-pyt-listings.py}{listing09}{listing10}{1}
  \codefor{code-for-listings.f}{listing10}{listing11}{1}
  \codefor{code-for-listings.f}{listing11}{listing12}{1}

  The $A$ term defined in equation \ref{eq:A} takes the following form:
  \codecpp{code-cpp-listings.hpp}{listing11}{listing12}{1}
  \codepyt{code-pyt-listings.py}{listing10}{listing11}{1}
  \codefor{code-for-listings.f}{listing12}{listing13}{1}

  The $B$ term defined in equation \ref{eq:B} takes the following form:
  \codecpp{code-cpp-listings.hpp}{listing12}{listing13}{1}
  \codepyt{code-pyt-listings.py}{listing11}{listing12}{1}
  \codefor{code-for-listings.f}{listing13}{listing14}{1}

  Equation \ref{eq:antidiff2} takes the following form:
  \codecpp{code-cpp-listings.hpp}{listing13}{listing14}{1}
  \codepyt{code-pyt-listings.py}{listing12}{listing13}{1}
  \codefor{code-for-listings.f}{listing14}{listing15}{1}

  Equation \ref{eq:antidiff1} takes the following form:
  \codecpp{code-cpp-listings.hpp}{listing14}{listing15}{1}
  \codepyt{code-pyt-listings.py}{listing13}{listing14}{1}
  \codefor{code-for-listings.f}{listing15}{listing16}{1}
  \codefor{code-for-listings.f}{listing16}{listing17}{1}

  The above listings conclude the formula-translation part of this paper.
  Implementation of a prototype MPDATA solver using the above code
    is presented in appendix~\ref{app:solver}.

  \section{Performance evaluation}\label{sec:perf}

  \subsection{Set up}

  The three introduced implementations of MPDATA were tested with the following set-ups 
    employing free and open-source tools:
  \begin{description}
    \item[C++:]{~
      \begin{itemize}
        \item{GCC g++ 4.8.0\footnote{\label{fnt:gcc-snapshot}GNU Compiler Collection packaged in the Debian's gcc-snapshot\_20130222-1} 
          and Blitz++ 0.10}
        \item{LLVM Clang 3.2 and Blitz 0.10}
      \end{itemize}
    }
    \item[Python:]{~
      \begin{itemize}
        \item{CPython 2.7.3 and NumPy 1.7}
        \item{PyPy 1.9.0 with built-in NumPy implementation}
      \end{itemize}
    }
    \item[Fortran:]{~
      \begin{itemize}
        \item{GCC gfortran 4.8.0\textsuperscript{\ref{fnt:gcc-snapshot}}}
      \end{itemize}
    }
  \end{description}
  The performance tests were run on a Debian and an Ubuntu GNU/Linux systems with the above-listed software obtained 
    via binary packages from the distributions' package repositories (most recent package versions at the time of writing).
  The tests were performed on two 64-bit machines equipped with
    an AMD Phenom\textsuperscript{\texttrademark} II X6 1055T (800 MHz)
    and an Intel\textsuperscript{\textregistered} Core\textsuperscript{\texttrademark}~i5-2467M (1.6 GHz)
    processors.

  For both C++ and Fortran the GCC compilers were invoked with the \prog{-Ofast} and the 
    \prog{-march=native} options.
  The Clang compiler was invoked with the \prog{-O3}, the \prog{-mllvm -vectorize}, the \prog{-ffast-math} 
    and the \prog{-march=native} options\footnote{Clang 3.4 enables vectorisation by default and supports the \prog{-Ofast} option}.
  The CPython interpreter was invoked with the \prog{-OO} option.

  In addition to the standard Python implementation CPython,
    the Python code was tested with PyPy.
  PyPy is an alternative implementation of Python featuring a just-in-time compiler. 
  PyPy includes an experimental partial re-implementation of NumPy that compiles NumPy expressions into native assembler.
  Thanks to employment of lazy evaluation of array expressions (cf. Sect. \ref{sec:array})
    PyPy allows to eliminate the use of temporary matrices for storing intermediate results,
    and to perform multiple operations on the arrays within a single array index traversal
    \footnote{Lazy evaluation available in PyPy 1.9 has been temporarily removed from PyPy during a refactoring of
      the code. It'll be reinstantiated in the codebase as soon as possible, but past PyPy 2.0 release}.
  Consequently, PyPy allows to overcome the same performance-limiting factors as those addressed by Blitz++, although 
    the underlying mechanisms are different.
  In contrast to other solutions for improving performance of NumPy-based codes such as
    Cython\footnote{see \url{http://cython.org}}, 
    numexpr\footnote{see \url{http://code.google.com/p/numexpr/}} or 
    Numba\footnote{see \url{http://numba.pydata.org/}}, 
    PyPy does not require any modifications to the code.
  Thus, PyPy may serve as a drop-in replacement for CPython ready to be used with 
    previously-developed codes.

  The same set of tests was run with all four set-ups.
  Each test set consisted of 16 program runs.
  The test programs are analogous to the example code presented in appendix~\ref{app:usage}.
  The tests were run with different grid sizes ranging from 64$\times$64 to 2048$\times$2048.
  The Gaussian impulse was advected for $nt=2^{24}/(nx\cdot ny)$ timesteps ($2^{24}$ chosen arbitrarily), 
    in order to assure comparable timing accuracy for all grid sizes.
  Three MPDATA iterations were used (i.e. two corrective steps).
  The tests were run multiple times; program start-up, data loading, and output verification times were
    subtracted from the reported values (see caption of Figure~\ref{fig:cpu-eyrie} for details).

>>>>

  \begin{tikzpicture}
    \coordinate (Origin)   at (0,0);
    \coordinate (XAxisMin) at (-2.5,0);
    \coordinate (XAxisMax) at (3,0);
    \coordinate (YAxisMin) at (0,-2.5);
    \coordinate (YAxisMax) at (0,3);
    \draw [thin, gray,-latex] (XAxisMin) -- (XAxisMax);% Draw x axis
    \draw [thin, gray,-latex] (YAxisMin) -- (YAxisMax);% Draw y axis

@@

    \draw [ultra thick,-latex,red] (0,.8) -- (0,1.2) node [above] {};
    \draw [ultra thick,-latex,red] (0,-1.2) -- (0,-.8) node [below right] {$C^{[y]}_{[i,j\mhlf]}$};
    \draw [ultra thick,-latex,red] (-2,.8) -- (-2,1.2) node [above] {};
    \draw [ultra thick,-latex,red] (-2,-1.2) -- (-2,-.8) node [above] {};
    \draw [ultra thick,-latex,red] (2,.8) -- (2,1.2) node [above] {};
    \draw [ultra thick,-latex,red] (2,-1.2) -- (2,-.8) node [above] {};
  \end{tikzpicture}
